 

### 2021-08-04 vertical-thermal-review.Rmd
abstract edits

Rising temperatures are having profound effects on forests; however, it is not well understood how responses vary vertically across forest strata. Here, we review how environmental conditions, leaf temperatures,  and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function. In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and are therefore at risk of elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.  However, foliar traits relevant to shaping $T_{leaf}$ also vary strongly across height or light gradients, partially offsetting the propensity towards elevated $T_{leaf}$ in the upper canopy. Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher metabolism and growth at both individual and ecosystem levels. Yet, differences in thermal sensitivity and damage thresholds are modest. The implications for tree performance are mixed: whereas larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may be stifled under hot, humid conditions. As warming continues, large canopy trees will be particularly vulnerable to drought, whereas understory trees will suffer most when the buffering provided by large trees is lost.

@@

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

>>>>

Rising temperatures are profoundly influencing forests on many scales with potentially strong variation vertically across forest strata. Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function. In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and can experience elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.  However, foliar traits that also vary strongly across height or light gradients can partially mitigate the elevation of $T_{leaf}$ in the upper canopy. Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels. Yet, differences in thermal sensitivity and damage thresholds across the gradient are modest. Under climate warming and increased drought, larger trees can face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.

@@

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
